Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-06-21 paper/submit-aas/referee-response.txt

* Comments from the AAS statistics editor:

> In Fig 2, the authors choose a parabolic model for the GD-1 stream with
> (arbitrary?) lower/upper boundaries. This band clearly includes some
> background stars, and may miss some stream stars around \phi_1=-30d. It would
> be nice to identify stream members by some statistical technique, either in
> the 2D sky or 4D sky+kinematics space.

> I suggest the authors consult the excellent summary of modern `spatial point
> processes' methodology with cookbook to the comprehensive 'spatstat' R/CRAN
> package, by A.Baddeley et al. 'Spatial Point Patterns: Methodology and
> Applications with R' (2015). For example, I suspect a good list of stream
> members could emerge from a Voronoi tessellation with truncation of tile
> areas. Other possibilities include: nearest neighbors within a distance
> threshold, nonparametric density estimation, Allard-Fraley feature detection
> (MLE on Voronoi tessellation), and Byers-Raftery feature detection (MLE
> mixture model on k-NN distribution).

The polyomial model (in phi1, phi2) is fit to the stream ridge points, then shifted up and down to define a selection region (as described in Section 3.1).

@@

We agree that, to characterize the stream, it would be better to build a probabilistic model of the stream in all available data dimensions.
However, to be useful for future work, this would require simultaneously modeling the phase-space density of the stream so that the variations with distance and distribution of proper motions could be taken into account in a robust way.
As the focus of this work is mainly to point out unexpected, off-track features that appear related to the stream, we defer this more sophisticated modeling to future work.

=========================================
For the journal editor:

We have addressed each of the referee's comments in the attached response, but in doing so we have added a new figure that we believe improves the paper and directly addresses one of the referee's points. We realize that now the word count is slightly over the limit for ApJ letters (we count XX words). However, we do still feel that this result and paper are in the spirit of the Letters and would like to keep it in consideration for this journal. Please let us know if we should work on trimming the text, or else please advise. Thanks!

### 2018-06-21 paper/gd1.tex
minor change

\changes{
To quantify the significance of these features, we construct a statistical model of the linear density of the filtered sky positions as a function of $\phi_2$ around each of the features and at a control region.
We select three equal ranges of $\phi_1$ centered on the spur, $\phi_1 \in (-36, -30)^\circ$, the blob, $\phi_1 \in (-18, -12)^\circ$, and in between (as a control field), $\phi_1 \in (-51, -45)^\circ$, and model the one-dimensional density in $\phi_2$ in each of these regions. % to see if the number of stars at the location of each feature is consistent with the background level.
We use a four-component mixture model to represent the density consisting of a uniform background over the range $\phi_2 \in (-10, 5)^\circ$, a two-component Gaussian mixture to represent the main stream with mean $\mu_{\textrm{s}}$ and variances $\sigma_{\textrm{s}, 1}^2$ and $\sigma_{\textrm{s}, 2}^2$, and a single Gaussian for each feature with mean and variance $(\mu_{\textrm{f}}, \sigma_{\textrm{f}}^2)$.
The full density model given all parameters $\bs{\theta} = (\alpha_{\textrm{bg}}, \alpha_{\textrm{s}}, \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}, \sigma_{\textrm{s}, 2}, \mu_{\textrm{f}}, \sigma_{\textrm{f}})$ is
\begin{equation}
    \begin{split}
    p(\phi_2 \given \bs{\theta}) &=
            \alpha_{\textrm{bg}} \, \mathcal{U}(-10, 5) \\ & \quad +
            \alpha_{\textrm{s}, 1} \, \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}) +
            \alpha_{\textrm{s}, 2} \, \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 2}) \\ & \quad +
            \alpha_{\textrm{f}} \,
                \mathcal{N}(\phi_2 \given \mu_{\textrm{f}}, \sigma_{\textrm{f}})
    \end{split}
\end{equation}
where $\alpha_{\textrm{bg}}$ is the fraction of background stars in the field, $\alpha_{\textrm{s}} = \alpha_{\textrm{s}, 1} + \alpha_{\textrm{s}, 2}$ is the fraction of stars associated with the mean stream track, and $\alpha_{\textrm{f}} = (1 - \alpha_{\textrm{bg}} - \alpha_{\textrm{s}})$ is the fraction of stars associated with either the spur, the blob, or at the location of the spur/blob in the control field.
% and ignore uncertainties on the sky position $\phi_2$.
}
\changes{
We generate posterior samples in the parameters of this model in each $\phi_1$ range by specifying uniform priors over all parameters except the variances, for which we use priors that are uniform in the logarithm of the variance.
We use an ensemble Markov Chain Monte Carlo (MCMC) sampler (\citealt{Foreman-Mackey:2013}) to generate posterior samples and run the sampler with 64 walkers for a total of 4096 steps, discarding the first 2048 samples as ``burn-in.''
\figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) with maximum a posteriori inferred density over-plotted (black curve) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $\phi_1$ range.
The 15th and 85th percentile values of the feature amplitude in different fields are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
The posterior samples for the amplitude of the feature mixture component are consistent with 0 in the control field, marginally inconsistent with 0 in the blob field, and significantly inconsistent with 0 in the spur field.
% In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.
% In each of the $\phi_1$ regions, the 15th and 85th percentile values of the feature amplitude, $\alpha_{\textrm{f}}$, are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
}

>>>>

\changes{
To quantify the significance of these features, we construct a statistical model of the linear density of the filtered sky positions as a function of $\phi_2$ around each of the features and at a control region.
We select three equal ranges of $\phi_1$ centered on the spur, $\phi_1 \in (-36, -30)^\circ$, the blob, $\phi_1 \in (-18, -12)^\circ$, and in between (as a control field), $\phi_1 \in (-51, -45)^\circ$, and model the one-dimensional density in $\phi_2$ in each of these regions. % to see if the number of stars at the location of each feature is consistent with the background level.
We use a four-component mixture model to represent the density consisting of a uniform background over the range $\phi_2 \in (-10, 5)^\circ$, a two-component Gaussian mixture to represent the main stream with mean $\mu_{\textrm{s}}$ and variances $\sigma_{\textrm{s}, 1}^2$ and $\sigma_{\textrm{s}, 2}^2$, and a single Gaussian for each feature with mean and variance $(\mu_{\textrm{f}}, \sigma_{\textrm{f}}^2)$.
The full density model given all parameters $\bs{\theta} = (\alpha_{\textrm{bg}}, \alpha_{\textrm{s}}, \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}, \sigma_{\textrm{s}, 2}, \mu_{\textrm{f}}, \sigma_{\textrm{f}})$ is
\begin{equation}
    \begin{split}
    p(\phi_2 \given \bs{\theta}) &=
            \alpha_{\textrm{bg}} \, \mathcal{U}(-10, 5) \\ & \quad +
            \alpha_{\textrm{s}, 1} \, \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}) +
            \alpha_{\textrm{s}, 2} \, \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 2}) \\ & \quad +
            \alpha_{\textrm{f}} \,
                \mathcal{N}(\phi_2 \given \mu_{\textrm{f}}, \sigma_{\textrm{f}})
    \end{split}
\end{equation}
where $\alpha_{\textrm{bg}}$ is the fraction of background stars in the field, $\alpha_{\textrm{s}} = \alpha_{\textrm{s}, 1} + \alpha_{\textrm{s}, 2}$ is the fraction of stars associated with the mean stream track, and $\alpha_{\textrm{f}} = (1 - \alpha_{\textrm{bg}} - \alpha_{\textrm{s}})$ is the fraction of stars associated with either the spur, the blob, or at the location of the spur/blob in the control field.
% and ignore uncertainties on the sky position $\phi_2$.
}
\changes{
We generate posterior samples in the parameters of this model in each $\phi_1$ range by specifying uniform priors over all parameters except the variances, for which we use priors that are uniform in the logarithm of the variance.
We use an ensemble Markov Chain Monte Carlo (MCMC) sampler (\citealt{Foreman-Mackey:2013}) to generate posterior samples and run the sampler with 64 walkers for a total of 4096 steps, discarding the first 2048 samples as ``burn-in.''
\figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) with maximum a posteriori inferred density over-plotted (black curve) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $\phi_1$ range.
The 15th and 85th percentile values of the feature amplitude in different fields are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
The posterior samples for the amplitude of each feature mixture component are consistent with zero in the control field, marginally significant in the blob field, and significantly inconsistent with the background in the spur field.
% In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.
% In each of the $\phi_1$ regions, the 15th and 85th percentile values of the feature amplitude, $\alpha_{\textrm{f}}$, are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
}
